---
title: "Bollt_FinalProject_Thermocline"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The crux of this project is to compare how the thermoclines on each of the 9 Wisconsin lakes in the NTL-LTER dataset have moved over the course of the 35 years of data. I am interested in this question because I have an interest in coldwater fisheries that rely on thermoclines to survive the summer weather. Perhaps climate change is affecting where the thermocline sets up, and therefore shrinking the available summer habitat of trout.
```{r package and data load; theme}
#Setting up session and loading all packages I think I might need
getwd()
library(tidyverse)
library(lubridate)
library(ggplot2)
library(multcompView)
library(nlme)
library(lsmeans)

#Reading in the raw lake dataset
NTLR_raw <- read.csv("V:/ENV_872_Project_Directory/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
View(NTLR_raw)

#setting a theme for plots
mytheme <- mytheme <- theme_classic(base_size = 12.2696)+
  theme(axis.text = element_text(color = "Blue"),
  legend.position = "top")
theme_set(mytheme)
```
Next steps (document everything with comments as you work):
Change the date column to dates
Filter to generate a dataset for each of the nine lakes.
Determine how to set a thermocline benchmark (July average?)
Determine what data to compare to over time (see Kateri's comment)
Run that comparison for each lake
Figure out how to compare each of the lakes to eachother to look for overall trends - maybe correlated with maximum depth
-maybe compare average temperature and temperature range?
-Mann-Kendall to look for trends between lakes controlling for seasonality (o2 and temperature)
-Look at each lake individually for both o2 and temperature to look for trends
Make a bunch of graphs
Write the report (you might have to knit it to Word for your own sanity)
Go fishing

Change the date column to dates
```{r date as date}
#Changing the date column to read as a date
class(NTLR_raw$sampledate)
NTLR_raw$sampledate <- as.Date(NTLR_raw$sampledate, format="%m/%d/%y")
class(NTLR_raw$sampledate)
```

Filter to generate a dataset for each of the nine lakes.
```{r filter}
unique(NTLR_raw$lakename)  # This data summary gives me the names of the nine lakes.
Paullake_raw <- NTLR_raw %>%
  filter(lakename == "Paul Lake")
Peterlake_raw <- NTLR_raw %>%
  filter(lakename == "Peter Lake")
Tuesdaylake_raw <- NTLR_raw %>%
  filter(lakename == "Tuesday Lake")
Eastlonglake_raw <- NTLR_raw %>%
  filter(lakename == "East Long Lake")
Westlonglake_raw <- NTLR_raw %>%
  filter(lakename == "West Long Lake")
Centrallonglake_raw <- NTLR_raw %>%
  filter(lakename == "Central Long Lake")
Hummingbirdlake_raw <- NTLR_raw %>%
  filter(lakename == "Hummingbird Lake")
Cramptonlake_raw <- NTLR_raw %>%
  filter(lakename == "Crampton Lake")
Wardlake_raw <- NTLR_raw %>%
  filter(lakename == "Ward Lake")
```
Determine how to set a thermocline benchmark (July average?)
"Keith, great idea. You might run into limitations answering this question due to sparse depth measurements and/or not enough temporal resolution. In that case, I might suggest a few options: (1) choose certain depths and compare along time series (i.e., hold depth constant and compare across time), (2) pay attention to the depth_id, which corresponds to a certain percent of surface irradiance (see NTL-LTER website for more info). Light could be used in addition to or separately from temperature data, (3) Analyze O2 in addition to temperature - this will give you an indication about development of hypoxia (when and at what depth) and how that might change over time. So, lots of places to pivot if you don't have enough resolution to answer your question at face value.''

Determine how to set a thermocline benchmark:
It looks like there is not enough temporal resolution to measure close-to-continuous change in thermocline depth over the course of a given season. Likewise, there is not enough close-to-continous depth measurements taken at each lake, nor is there consistant data taken below 10 meters of depth. As such, I will compare what is occurring at 7 meters depth at each lake over time. 7 meters, I know from my experience as a fisherman, is on the shallow end of where a thermocline sets up in a northern US lake. Therefore, evaluating what is happening at 7m depth in each lake will give a good idea what sorts of conditions trout are dealing with in these lakes in the summmer. I expect there to be more variation by year at 7 meters depth than at, say, 10 meters depth. Looking at 7 meters will give a better picture of whether the thermocline in these lakes is changing over time. Of course, this is more academic than practical if trout do not live in these lakes.

I will look at two indicators of thermocline establishment at 7 meters: temperature and dissolved oxygen content. I would expect water at or below the thermocline to have low temperatures and hypoxic conditions (low oxygen levels).

My research question, then, is as follows:

How have temperature and oxygen conditions changed at 7 meters depth in a series of Wisconsin Lakes? Is climate change affecting where the thermocline sets up in these lakes?
```{r processing each lake}
#filtering for just the data at 7m depths in summer (June 20-September 21)
Paullake_processed <- Paullake_raw %>%
  filter(depth == 7, daynum %in% 172:264) %>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()
dim(Paullake_processed)  
#Enough datapoints to analyse this lake.

Peterlake_processed <- Peterlake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()
dim(Peterlake_processed) 
#Enough datapoints to analyse this lake.

Tuesdaylake_processed <- Tuesdaylake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()
dim(Tuesdaylake_processed) #Enough datapoints to analyse this lake.

Eastlonglake_processed <- Eastlonglake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()
dim(Eastlonglake_processed) 
#Enough datapoints to analyse this lake.

Westlonglake_processed <- Westlonglake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()
dim(Westlonglake_processed) 
#Enough datapoints to analyse this lake.

Centrallonglake_processed <- Centrallonglake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()  
dim(Centrallonglake_processed) 
#There are 0 data points taken on Central Long Lake below 4 meters. Unfortunately, I am going to have to eliminate Central Long Lake from my analysis.

Hummingbirdlake_processed <- Hummingbirdlake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()  
dim(Hummingbirdlake_processed) 
#Hummingbird lake only has one datapoint at 7 meters. Unfortunately, I am going to have to eliminate Hummingbird Lake from my analysis.

Cramptonlake_processed <- Cramptonlake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude()
dim(Cramptonlake_processed) 
#There are only 34 datapoints, not enough datapoints to analyse this lake.

Wardlake_processed <- Wardlake_raw %>%
  filter(depth == 7, daynum %in% 172:264)%>%
  select(lakename:dissolvedOxygen) %>%
  mutate(Week = week(sampledate)) %>%
  na.exclude() 
dim(Wardlake_processed) 
#Ward lake only has 11 datapoints at 7 meters. Unfortunately, I am going to have to eliminate Ward Lake from my analysis.
```
After deciding on what question I wanted to answer and what data I wanted to use to answer my question, I unfortunately had to eliminate three lakes from my analysis. I still have five lakes to perform my analysis on.

```{r combining processd data}
#Combining the processed lake data into one file for later analysis
NTL_processed <- rbind(Paullake_processed, Peterlake_processed, Tuesdaylake_processed,
                          Eastlonglake_processed, Westlonglake_processed)
View(NTL_processed)
unique(NTL_processed$lakename)
dim(NTL_processed)
head(NTL_processed)
```

A little data visualization before my data analysis.
```{r visualization, include=FALSE}
Alllakes_boxplot <- ggplot(NTL_processed, aes(x = lakename, y = temperature_C, color = lakename))+
  geom_boxplot()+
  labs(x = "Lake Name", y = "Temperature (Celsius)", color = "Lake Name")
Alllakes_boxplot
#This graph tells us about the temperature data distribution for each lake at 7m depth, and tells us which lakes are coldest in the summer. 

Paultemp_visualization <- ggplot(Paullake_processed, aes(x = sampledate, y = temperature_C, color = temperature_C))+
  geom_point()+
  labs(x = "Date", y = "Temperature in Celsius", color = "Temperature (Degrees Celsius)")+
  scale_color_gradient(low = "dark blue", high = "light blue")
Paultemp_visualization
#Not really much of a trend in temperature over time.

Paulo2_visualization <- ggplot(Paullake_processed, aes(x = sampledate, y = dissolvedOxygen, color = dissolvedOxygen))+
  geom_point()+
  labs(x = "Date", y = "Percent Dissolved O2", color = "Dissolved O2")+
  scale_color_gradient(low = "red", high = "white")
Paulo2_visualization
#I can't really see a trend in dissolved oxygen

Alllakes_visualization <- ggplot(NTL_processed, aes(x = sampledate, y = temperature_C, color = lakename))+
  geom_point()+
  labs(x = "Date", y = "Temperature (Celsius)", color = "Lake Name")
Alllakes_visualization
#Kind of hard to see a trend here. This graph has too much data at once and different lakes have different temporal ranges.
```

Now that I have visualized my data, it is time to start running statistical tests on it in order to answer my research question. I am going to look at each of the nine lakes individually, and then look at all nine lakes at once. I am interested in two parameters at 7 meters depth, temperature and dissolved oxygen content. I will run a repeated measures ANOVA on the combined processed dataset, which contains data from five lakes. I will first run the test on temperature and then on dissolved oxygen. This test takes into account autocorrelation within a given year and within a given lake. 

```{r repeated measures ANOVA temperature }

#Accounting for autocorrelation
Alllakestemptest.auto <- lme(data = NTL_processed, 
                     temperature_C ~ sampledate * lakename, #fixed effects portion of model: response variable, being predicted based off of sampledate and lakename
                     random = ~1|Week)  # this is the random effect portion of the model
Alllakestemptest.auto
# we care about the Stddeviation between each week
ACF(Alllakestemptest.auto)
# we care about the lag of 1's value. This tells us how much temperature is autocorrelated within a given year (it's about 23%)

#running the ANOVA
Alllakestemptest.mixed <- lme(data = NTL_processed,
                     temperature_C ~ sampledate * lakename, 
                     random = ~1|Week,
                     correlation = corAR1(form = ~ sampledate/lakename|Week, value = 0.2323), #correlation from previous model, sampledate/lakename because the model can only do one lake at a time
                     method = "REML")
summary(Alllakestemptest.mixed)
#There is not a significant trend among all of the lakes at 7m.
```
The results from our mixed effects test demonstrates an important finding. Observe that the p-value for sampledate is 0.94. This means that we can reject the null hypothesis that there is a significant correlation between date and temperature at 7 meters. This does not supports the idea that the temperature at 7 meters in these lakes is changing. I am going to run the test on individual lakes later.

Now, I will run the same test on the five lakes, substituting oxygen for temperature. 

```{r repeated measures ANOVA oxygen}
#oxygen may be autocorrelated with temperature
#copy from above, and plug in the oxygen data
#Accounting for autocorrelation
Alllakesoxygentest.auto <- lme(data = NTL_processed, 
                     dissolvedOxygen ~ sampledate * lakename, #fixed effects portion of model: response variable, being predicted based off of sampledate and lakename
                     random = ~1|Week)  # this is the random effect portion of the model
Alllakesoxygentest.auto
# we care about the Stddeviation between each week
ACF(Alllakesoxygentest.auto)
# we care about the lag of 1's value. This tells us how much temperature is autocorrelated within a given year (it's about 6%)

#running the ANOVA
Alllakesoxygentest.mixed <- lme(data = NTL_processed,
                     dissolvedOxygen ~ sampledate * lakename, 
                     random = ~1|Week,
                     correlation = corAR1(form = ~ sampledate/lakename|Week, value = 0.0645), #correlation from previous model, sampledate/lakename because the model can only do one lake at a time
                     method = "REML")
summary(Alllakesoxygentest.mixed)
#There is not a significant trend among all of the lakes at 7m.
```
The results from our mixed effects test demonstrates an important finding. Observe that the p-value for sampledate is 0.43. This means that we can reject the null hypothesis that there is a significant correlation between date and dissolved oxygen content at 7 meters. This does not supports the idea that the dissolved oxygen at 7 meters in these lakes is changing. I am going to run the test on individual lakes later.
```{r ANOVA temperature by lake}
Paultemptest.auto <- lme(data = Paullake_processed, 
                     temperature_C ~ sampledate, 
                     random = ~1|Week)  
Paultemptest.auto
ACF(Paultemptest.auto)  # 0.4057 autocorrelation
Paultemptest.mixed <- lme(data = Paullake_processed,
                     temperature_C ~ sampledate, 
                     random = ~1|Week,
                     correlation = corAR1(form = ~ sampledate|Week, value = 0.4057),
                     method = "REML")
summary(Paultemptest.mixed)
#Paul Lake had a p-value less than 0.05 and a slope of 0.000045/day.

Petertemptest.auto <- lme(data = Peterlake_processed, 
                     temperature_C ~ sampledate, 
                     random = ~1|Week)  
Petertemptest.auto
ACF(Petertemptest.auto)  # 0.3383 autocorrelation
Petertemptest.mixed <- lme(data = Peterlake_processed,
                     temperature_C ~ sampledate, 
                     random = ~1|Week,
                     correlation = corAR1(form = ~ sampledate|Week, value = 0.3383),
                     method = "REML")
summary(Petertemptest.mixed)
#Peter Lake had a p-value less than 0.05 and a slope of -0.0001/day.
```

